Bow shocks, bow waves, and dust waves. II. Beyond the rip point

### 2019-03-08 papers/Papers1234/sec-strong-coupling.tex
Jane's Paper I corrections, up to end of sec 2.3

\begin{figure*}
  \includegraphics[width=\linewidth]{figs/zones-v-n-plane}
  \caption{Bow regimes in parameter space (\(v, n\)) of the external
    stream for main-sequence OB stars of different masses:
    (a)~\SI{10}{M_\odot}, (b)~\SI{20}{M_\odot}, (c)~\SI{40}{M_\odot}.  In all
    cases, \(\kappa = \SI{600}{cm^2.g^{-1}}\) and efficient gas-grain
    coupling is assumed. Solid black lines of varying width show the
    bow size (star-apex separation, \(R_0\)), while gray shading shows
    the radiation bow wave regime, with lower border
    \(\tau = \eta\wind\) and upper border \(\tau = 1\), where
    \(\tau = 2 \kappa \rho R_0\) is the optical depth through the bow.  For bows
    above the red solid line, the ionization front is trapped inside
    the bow (\S~\ref{sec:trapp-ioniz-front}).  Blue lines delineate
    different cooling regimes (\S~\ref{sec:radi-cool-lengths}).  Above
    the thin blue line (\(d_{\text{cool}} = h_0\)), the bow shock
    radiates efficiently, forming a thin shocked shell.  Below the
    thick blue line (\(d_{\text{cool}} = R_0\)), the bow shock is
    essentially non-radiative.}
  \label{fig:zones-v-n-plane}
\end{figure*}

@@

Although dust grains typically constitute only a small fraction
\(Z\grain \sim 0.01\) of the mass of the external stream, they
nevertheless dominate the broad-band opacity at FUV, optical and IR
wavelengths if they are present.\footnote{%
  At EUV wavelengths (\(\lambda < \SI{912}{\angstrom}\)), gas opacity
  dominates if the hydrogen neutral fraction is larger than
  \(\approx 0.001\), see discussion of ionization front trapping
  below.} %
The strong coupling assumption means that all the radiative forces
applied to the dust grains are directly felt by the gas also.

@@

The internal pressure is the sum of wind ram pressure and the
effective radiation pressure that acts on the bow shell.  The
radiative momentum loss rate of the star is \(L/c\) and the wind
momentum loss rate can be expressed as
\begin{equation}
  \label{eq:wind-efficiency}
  \dot{M} V = \eta\wind L / c \ , 
\end{equation}
where \(\eta\wind\) is the momentum efficiency of the wind, which is typically
\(< 1\) \citep{Lamers:1999b}. If the optical depth is very large, then
all of the stellar radiative momentum, emitted with rate \(L/c\), is
trapped by the bow shell.  In the single scattering limit,\footnote{%
  Although it may seem inconsistent to assume single scattering in the
  case of high optical depths, this is defensible for the following
  reasons. (1)~The grain albedo is not that high (typically
  \(\sim 0.5\) at ultraviolet through optical wavelengths). (2)~The
  scattered radiation field is more isotropic than the stellar field,
  leading to cancellation in the radiative flux. (3)~Absorbed
  radiation is re-emitted at infrared wavelengths, where the dust
  opacity is very much lower.} %
and temporarily neglecting the wind for clarity of exposition, then
pressure balance at the bow apex, a distance \(R_0\) along the
symmetry axis from the star is given by
\begin{equation}
  \label{eq:rad-press-balance-thick}
  \frac{L}{4 \pi c R_0^2} = \rho v^2 \ ,
\end{equation}
which yields a fiducial bow shock radius in this optically thick,
radiation-only limit as
\begin{equation}
  \label{eq:Rstar}
  R_* = \left(\frac{L}{4\pi c \rho v^2}\right)^{1/2} \ .
\end{equation}

We now consider the opposite, optically thin limit.  If the total
opacity (gas plus dust) per total mass (gas plus dust) is \(\kappa\) (with
units of \si{cm^2.g^{-1}}), then the radiative acceleration is
\begin{equation}
  \label{eq:rad-accel}
  a_{\text{rad}} = \frac{\kappa L}{4 \pi c R^2} \ .
\end{equation}
Therefore, an incoming stream with initial velocity, \(v_\infty\), can be
brought to rest by radiation alone
% \footnote{%
%   For simplicity, we here ignore the effects of gravity, which are
%   important for low ratios of \(\kappa L / M\), see
%   \S~\ref{sec:effects-gravity}.  We also ignore pressure gradients and
%   shocks, which are important as the velocity approaches the sound
%   speed, \(\sound\) (in \S~XX below, we show that the resultant
%   corrections to \(R_0\) are of order \(\sound / v_\infty\)).} %
at a distance \(R\starstar\) where
\begin{equation}
  \label{eq:rad-poten}
  \int_{R\starstar}^\infty a_{\text{rad}} \, dr = \tfrac12 v_\infty^2 \ , 
\end{equation}
yielding
\begin{equation}
  \label{eq:rad:R0}
  R\starstar = \frac{\kappa L}{2\pi c v_\infty^2} \ .
\end{equation}
On the other hand, we can also argue as in the optically thick case
above by approximating the bow shell as a surface, and balancing
stellar radiation pressure against the ram pressure of the incoming
stream.  The important difference when the shell is not optically
thick is that only a fraction \(1 - e^{-\tau}\) of the radiative momentum
is absorbed by the bow, so that
equation~\eqref{eq:rad-press-balance-thick} is replaced with
\begin{equation}
  \label{eq:rad-press-balance-tau}
  \frac{L (1 - e^{-\tau})}{4 \pi c R_0^2} = \rho v^2 \ .
\end{equation}
In the optically thin limit, \(1 - e^{-\tau} \approx \tau\), so these two
descriptions can be seen to agree (\(R_0 \to R\starstar\)) so long as
\begin{equation}
  \label{eq:tau-thin}
  \tau = 2 \kappa \rho R_0 \ ,
\end{equation}
which we will assume to hold generally.

Then, defining a fiducial optical depth,
\begin{equation}
  \label{eq:tau-star}
  \tau_* = \rho \kappa R_* \ ,
\end{equation}
and now reinstating the stellar wind ram pressure term
% \footnote{%
%   We implicitly assume that the interaction of the stellar wind with
%   the external stream can always be treated in the continuum limit.
%   This will be true if either the collisional mean free path or the
%   ion Larmor radius is much smaller than \(R_0\), which is almost
%   always the case.} %
from equation~\eqref{eq:wind-efficiency}, we find that the general bow
radius can be written in terms of the fiducial radius as
\begin{equation}
  \label{eq:R0-definition}
  R_0 = x R_* \ ,
\end{equation}
where \(x\) is the solution of
\begin{equation}
  \label{eq:rad-full-x}
  x^2 - \bigl(1 - e^{-2 \tau_* x} \bigr) - \eta\wind = 0 \ .
\end{equation}
Since this is a transcendental equation, \(x\) must be found
numerically, but we can write explicit expressions for three limiting
cases:
\begin{equation}
  \label{eq:x-cases}
  x \approx
  \begin{cases}
    \text{if \(\tau_* \gg 1\):} & (1 + \eta\wind)^{1/2}  \\
    \text{if \(\tau_*^2 \ll 1\):} & \tau_* + \bigl( \tau_*^2 + \eta\wind \bigr)^{1/2} \approx
    \begin{cases}
      \text{if \(\tau_*^2 \gg \eta\wind\):} & 2 \tau_*  \\
      \text{if \(\tau_*^2 \ll \eta\wind\):} & \eta\wind^{1/2} 
    \end{cases}
  \end{cases}
\end{equation}
The first case, \(x \approx (1 + \eta\wind)^{1/2}\), corresponds to a
\textit{radiation bow shock} (RBS); the second case,
\(x \approx 2 \tau_* \), corresponds to a \textit{radiation bow wave} (RBW);
and the third case, \(x \approx \eta\wind^{1/2}\), corresponds to a
\textit{wind bow shock} (WBS).  The two bow shock cases are similar in
that the external stream is oblivious to the presence of the star
until it suddenly hits the bow shock shell, differing only in whether
it is radiation or wind that is providing the internal pressure.  In
the intermediate bow wave case, on the other hand, the external stream
is gradually decelerated by absorption of photons as it approaches the
bow.\footnote{A shock can still form in this case, but shocked
  material constitutes only a small fraction of the total column
  density of the shell.}

@@

Figure~\ref{fig:M-supergiant} shows results for a cool M-type
super-giant star with stellar parameters inspired by Betelgeuse
(\chemalpha~Orionis), as listed in Table~\ref{tab:stars}.  Unlike the
UV-dominated spectrum of the hot stars, this star emits predominantly
in the near-infrared, where the dust extinction efficiency is lower,
so we adopt a lower opacity of \SI{60}{cm^2.g^{-1}}.  This has the
effect of shifting the radiation bow wave regime to higher densities:
\(n = \text{\numrange{1000}{30 000}}\, v_{10}^2\) in this case.

@@

In principle, gravitational attraction from the star, of mass \(M\),
will partially counteract the radiative acceleration.  This can be
accounted for by replacing \(L\) with an effective luminosity
\newcommand\Edd{\ensuremath{_{\text{E}}}}
\begin{equation}
  \label{eq:effective-luminosity}
  L_{\text{eff}} = L \bigl(1 - \Gamma\Edd^{\,-1}\bigr) \ ,
\end{equation}
in which \(\Gamma\Edd\) is the Eddington factor:
\begin{equation}
  \label{eq:eddington-factor}
  \Gamma\Edd = \frac{\kappa L}{4\pi c G M} = 458.5 \, \frac{\kappa_{600} L_4}{ M } \ ,
\end{equation}
where, in the last expression, \(M\) is measured in solar masses.  For
the stars in Table~\ref{tab:stars}, we find
\(\Gamma\Edd \approx \text{\numrange{30}{400}}\), so gravity can be safely
ignored.  When the optical depth of the bow is very large,
\(\tau > \ln\Gamma\Edd \sim 5\), gravity will exceed the radiation force in the
outer parts of the shell (see \citealt{Rodriguez-Ramirez:2016b}), but
it is generally too weak to effect the shell structure even in such a
case, see \S~\ref{sec:axial-structure-bow} below.

>>>>

\begin{figure*}
  \includegraphics[width=\linewidth]{figs/zones-v-n-plane}
  \caption{Bow regimes in parameter space (\(v, n\)) of the external
    stream for main-sequence OB stars of different masses:
    (a)~\SI{10}{M_\odot}, (b)~\SI{20}{M_\odot}, (c)~\SI{40}{M_\odot}.  In all
    cases, \(\kappa = \SI{600}{cm^2.g^{-1}}\) and efficient gas-grain
    coupling is assumed. Solid black lines of varying thickness show the
    bow size (star-apex separation, \(R_0\)), while gray shading shows
    the radiation bow wave regime, with lower border
    \(\tau = \eta\wind\) and upper border \(\tau = 1\), where
    \(\tau = 2 \kappa \rho R_0\) is the optical depth through the bow.  For bows
    above the red solid line, the ionization front is trapped inside
    the bow (\S~\ref{sec:trapp-ioniz-front}).  Blue lines delineate
    different cooling regimes (\S~\ref{sec:radi-cool-lengths}).  Above
    the thin blue line (\(d_{\text{cool}} = h_0\)), the bow shock
    radiates efficiently, forming a thin shocked shell.  Below the
    thick blue line (\(d_{\text{cool}} = R_0\)), the bow shock is
    essentially non-radiative.}
  \label{fig:zones-v-n-plane}
\end{figure*}

@@

Although dust grains typically constitute only a small fraction
\(Z\grain \sim 0.01\) of the mass of the external stream, they
nevertheless dominate the broad-band opacity at FUV, optical and IR
wavelengths if they are present. They also dominate at EUV wavelengths
(\(\lambda < \SI{912}{\angstrom}\)), if the hydrogen neutral fraction is
less than \(\approx 0.001\).  The strong coupling assumption means that all
the radiative forces applied to the dust grains are directly felt by
the gas also.

@@

The internal pressure is the sum of wind ram pressure and the
effective radiation pressure that acts on the bow shell.  The
radiative momentum loss rate of the star is \(L/c\) and the wind
momentum loss rate can be expressed as
\begin{equation}
  \label{eq:wind-efficiency}
  \dot{M} V = \eta\wind L / c \ , 
\end{equation}
where \(\eta\wind\) is the momentum efficiency of the wind, which is
\(< 1\) in all cases except WR stars \citep{Lamers:1999b}. If the
optical depth at UV/optical wavelengths is sufficiently large, then
all of the stellar radiative momentum, emitted with rate \(L/c\), is
trapped by the bow shell.  At the infrared wavelengths where the
absorbed radiation is re-emitted, the grain opacity is much lower, so
the bow is optically thin to this radiation.  Combined with the fact
that the UV single-scattering albedo is only \(\approx 0.5\), this means
that the single-scattering limit is approximately valid, and we adopt
it here.

We define a fiducial bow shock radius \(R_*\) in the optically thick,
radiation-only limit by balancing stellar radiation pressure against
stream ram pressure at the bow apex along the
symmetry axis from the star:
\begin{equation}
  \label{eq:rad-press-balance-thick}
  \frac{L}{4 \pi c R_*^2} = \rho v^2 \ ,
\end{equation}
which yields 
\begin{equation}
  \label{eq:Rstar}
  R_* = \left(\frac{L}{4\pi c \rho v^2}\right)^{1/2} \ .
\end{equation}

We now consider the opposite, optically thin limit.  If the total
opacity (gas plus dust) per total mass (gas plus dust) is \(\kappa\) (with
units of \si{cm^2.g^{-1}}), then the radiative acceleration is
\begin{equation}
  \label{eq:rad-accel}
  a_{\text{rad}} = \frac{\kappa L}{4 \pi c R^2} \ .
\end{equation}
Therefore, an incoming stream with initial velocity, \(v_\infty\), can be
brought to rest by radiation alone at a distance \(R\starstar\) where
\begin{equation}
  \label{eq:rad-poten}
  \int_{R\starstar}^\infty a_{\text{rad}} \, dr = \tfrac12 v_\infty^2 \ , 
\end{equation}
yielding
\begin{equation}
  \label{eq:rad:R0}
  R\starstar = \frac{\kappa L}{2\pi c v_\infty^2} \ .
\end{equation}
On the other hand, we can also argue as in the optically thick case
above by approximating the bow shell as a surface, and balancing
stellar radiation pressure against the ram pressure of the incoming
stream.  The important difference when the shell is not optically
thick is that only a fraction \(1 - e^{-\tau}\) of the radiative momentum
is absorbed by the bow, so that
equation~\eqref{eq:rad-press-balance-thick} is replaced with
\begin{equation}
  \label{eq:rad-press-balance-tau}
  \frac{L (1 - e^{-\tau})}{4 \pi c R_0^2} = \rho v^2 \ .
\end{equation}
In the optically thin limit, \(1 - e^{-\tau} \approx \tau\), so these two
descriptions can be seen to agree (\(R_0 \to R\starstar\)) so long as
\begin{equation}
  \label{eq:tau-thin}
  \tau = 2 \kappa \rho R_0 \ ,
\end{equation}
which we will assume to hold generally.

We define a fiducial optical depth,
\begin{equation}
  \label{eq:tau-star}
  \tau_* = \rho \kappa R_* \ .
\end{equation}
With these fiducial values established, we return to the combined
radiation plus wind scenario.  Adding the stellar wind ram pressure
term from equation~\eqref{eq:wind-efficiency} allows us to write the
general bow radius in terms of the fiducial radius as
\begin{equation}
  \label{eq:R0-definition}
  R_0 = x R_* \ ,
\end{equation}
where \(x\) is the solution of
\begin{equation}
  \label{eq:rad-full-x}
  x^2 - \bigl(1 - e^{-2 \tau_* x} \bigr) - \eta\wind = 0 \ .
\end{equation}
Since this is a transcendental equation, \(x\) must be found
numerically, but we can write explicit expressions for three limiting
cases:
\begin{equation}
  \label{eq:x-cases}
  x \approx
  \begin{cases}
    \text{if \(\tau_* \gg 1\):} & (1 + \eta\wind)^{1/2}  \\
    \text{if \(\tau_*^2 \ll 1\):} & \tau_* + \bigl( \tau_*^2 + \eta\wind \bigr)^{1/2} \approx
    \begin{cases}
      \text{if \(\tau_*^2 \gg \eta\wind\):} & 2 \tau_*  \\
      \text{if \(\tau_*^2 \ll \eta\wind\):} & \eta\wind^{1/2} 
    \end{cases}
  \end{cases}
\end{equation}
The first case, \(x \approx (1 + \eta\wind)^{1/2}\), corresponds to a
\textit{radiation bow shock} (RBS); the second case,
\(x \approx 2 \tau_* \), corresponds to a \textit{radiation bow wave} (RBW);
and the third case, \(x \approx \eta\wind^{1/2}\), corresponds to a
\textit{wind bow shock} (WBS).  The two bow shock cases are similar in
that the external stream is oblivious to the presence of the star
until it suddenly hits the bow shock shell, differing only in whether
it is radiation or wind that is providing the internal pressure.  In
the intermediate bow wave case, on the other hand, the external stream
is gradually decelerated by absorption of photons as it approaches the
bow.  We remark that a shock can still form in this case, but shocked
material constitutes only a fraction of the total column density of
the shell, as we will show in \S~\ref{sec:axial-structure-bow}.

@@

Figure~\ref{fig:M-supergiant} shows results for a cool M-type
super-giant star with stellar parameters inspired by Betelgeuse
(\chemalpha~Orionis), as listed in Table~\ref{tab:stars}.  Unlike the
UV-dominated spectrum of the hot stars, this star emits predominantly
in the near-infrared, where the dust extinction efficiency is lower \citep{Weingartner:2001a},
so we adopt a lower opacity of \SI{60}{cm^2.g^{-1}}.  This has the
effect of shifting the radiation bow wave regime to higher densities:
\(n = \text{\numrange{1000}{30 000}}\, v_{10}^2\) in this case.

@@

In principle, gravitational attraction from the star, of mass \(M\),
will partially counteract the radiative acceleration.  This can be
accounted for by replacing \(L\) with an effective luminosity
\newcommand\Edd{\ensuremath{_{\text{E}}}}
\begin{equation}
  \label{eq:effective-luminosity}
  L_{\text{eff}} = L \bigl(1 - \Gamma\Edd^{\,-1}\bigr) \ ,
\end{equation}
in which \(\Gamma\Edd\) is the Eddington factor:
\begin{equation}
  \label{eq:eddington-factor}
  \Gamma\Edd = \frac{\kappa L}{4\pi c G M} = 458.5 \, \frac{\kappa_{600} L_4}{ M } \ ,
\end{equation}
where, in the last expression, \(M\) is measured in solar masses.  For
the stars in Table~\ref{tab:stars}, we find
\(\Gamma\Edd \approx \text{\numrange{30}{400}}\), so gravity can be safely
ignored.  When the optical depth of the bow is very large,
\(\tau > \ln\Gamma\Edd \sim 5\), gravity will exceed the radiation force in the
outer parts of the shell (see \citealt{Rodriguez-Ramirez:2016b}), but
it is generally too weak to affect the shell structure even in such a
case, see \S~\ref{sec:axial-structure-bow} below.
